Modeling the effect of aneuploidy on cancer evolution

### 2023-05-30 EvolutionaryRescue1.tex
revised intro, results, figures - need to proceed from rescue prob

\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{tikz,pgfplots}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{xurl}

@@

@@

\author[1]{Remus Stana}
\author[2]{Uri Ben-David}
\author[3]{Daniel B. Weissman}
\author[1,*]{Yoav Ram}
\affil[1]{School of Zoology, Faculty of Life Sciences, Tel Aviv University}
\affil[2]{Department of Human Molecular Genetics and Biochemistry, Faculty of Medicine, Tel Aviv University}
\affil[3]{Department of Physics, Emory University}
\affil[*]{Corresponding author: yoav@yoavram.com}

@@

\paragraph{Aneuploidy in cancer.} Chromosomal instability (CIN) is the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy, where cells are characterized by structural changes of the chromosomes and copy number alterations \cite{schukken2018cin}.
Interestingly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy.
Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \cite{ben2020context}.

@@

The role of chromosomal instability (CIN) in the emergence of cancer has been studied extensively in the past decades \cite{michor2005can,christine2018understanding,nowak2002role,pavelka2010dr,komarova2003mutation,zhu2018cellular}.
One hypothesis is that CIN facilitates tumor genesis by accelerating the removal of tumor suppression genes (TSG) and subsequent appearance of cancer. The deletion of tumor suppression genes can happen in two ways: two point mutations deleting both alleles of the TSG (assuming a diploid genotype), or one point mutation and one chromosomal loss event.
Initial theoretical studies have shown that aneuploidy can have a significant role in the deletion of the the tumor suppressing genes when compared to two consecutive point mutations \cite{nowak2002role,komarova2003mutation,michor2005can,komarova2008selective}. % TODO mention if studies are theoretical , experimental, or clinical OK
However, when taking into account that the appearance of aneuploidy requires a mutation to trigger CIN, the probability that CIN precedes tumor genesis is highly unlikely.

\paragraph{Evolutionary rescue.} Populations adapted to a certain environment are vulnerable to environmental changes, which might cause extinction of the population. Examples of such environmental changes include climate change, invasive species or the onset of drug therapies. Adaptation is a race against time as the population size decreases in the new environment~\cite{tanaka2022surviving}. 
{\em Evolutionary rescue} is the process where the population acquires a trait that increases fitness in the new environment such that extinction is averted. It is mathematically equivalent to the problem of crossing of fitness valley \cite{weissman2009rate,weissman2010rate}.
There are three potential ways for a population to survive environmental change: migration to a new habitat similar to the one before the onset of environmental change \cite{cobbold2020should}; adaptation by phenotypic plasticity without genetic modification \cite{carja2019evolutionary,carja2017evolutionary,levien2021non}; and adaptation through genetic modifications, e.g., mutation \cite{uecker2014evolutionary,uecker2016role,uecker2011fixation}.

Models of evolutionary rescue usually assume that the fitness of the wildtype and mutant are independent of time. An exception is \cite{marrec2020adapt}, where the fitness of the wildtype and mutant are time dependent. Additionally, \cite{uecker2011fixation} investigated the probability of fixation of a beneficial mutation in a variable environment with arbitrary time-dependent selection coefficent and population size.
Most models focus on the probability that at least one mutation rescues the population. How multiple mutations contribute to the survival of the population is less explored. One exception is \cite{wilson2017soft} which showed that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute. 
Evolutionary rescue that requires two successive mutations has been investigated by \cite{martin2013probability} with the help of diffusion approximation.

@@

We follow the number of cancer cells that belong to three different types at time $t$: wildtype euploid, $w_t$; wildtype aneuploid, $a_t$; and mutant euploid, $m_t$. 
These cells grow and die with rates $\lambda_k$ and $\mu_k$ for $k=w, a, m$.
Wildtype cells become aneuploid at rate $u$. Aneuploid and wildtype cells mutate with rate $v$ (Figure \ref{figureAneuploidy}). 
Thus, the changes in the number of each cell type is described by
\begin{subequations}
\begin{flalign}
w_t&\rightarrow w_t+1:\quad \lambda_ww_t,\\
w_t&\rightarrow w_t-1:\quad \left(\mu_w+u+v\right)w_t,\\
a_t&\rightarrow a_t+1:\quad \lambda_aa_t+uw_t,\\
a_t&\rightarrow a_t-1:\quad \left(\mu_a+v\right)a_t,\\
m_t&\rightarrow m_t+1:\quad \lambda_am_t+va_t+vm_t,\\
m_t&\rightarrow m_t-1:\quad \mu_am_t.
\end{flalign}
\end{subequations}

The difference between growth and death is $\Delta_k = \lambda_k-\mu_k$.
We assume that $\Delta_w<0$ due to drug therapy (or some other stress), and that $\Delta_m$ due to resistance conferred by the mutant.
We analyze two cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are susceptible, $0>\Delta_a>\Delta_w$  \cite{brauner2016distinguishing}.

Here, \emph{evolutionary rescue} occurs when the aneuploid cells are able to either prevent (when $\Delta_a>0$) or delay (when $0>\Delta_a>\Delta_w$) the extinction of the cancer-cell population before the adaptive mutant cells appear and fixate in the populaiton.
To analyze evolutionary rescue in this model, we use the framework of \emph{multitype branching processes} \cite{rybnikov2021fitness,harris1963theory}. 
This allows us to find explicit expressions for the survival probability of the population--the probability that it does not become extinct due to stressful conditions, e.g., drug therapy.
Note that branching processes imply that lineages produced by cells from the initial population grow and die independently of each other.
This is a necessary simplification, but it is unrealistic, as cells usually compete for resources.
A more realistic model may include competition on limited resources between lineages, as well as spatial structure, which may play an important role in the development of cancer \cite{martens2011spatial}.

The survival probability of a population consisting initially of one wildtype cell satisfies the quadratic equation,
\begin{equation} \label{quadraticeqev1}
\begin{aligned}
1-p_w = & \frac{\mu_w}{\lambda_w+\mu_w+u+v} + 
		  \frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right) + \\
		  & \frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2 +
		  \frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right) ,
\end{aligned}
\end{equation}
where the $p_w$, $p_a$, and $p_m$ are the survival probabilities of a population consisting initially of one wildtype cell, one aneuploid cell, and one mutant cell, respectively.

The smallest solution of the quadratic equation \cref{quadraticeqev1} gives the survival probability \cite{uecker2015adaptive}, which still depends on both $p_a$ and $p_m$,
\begin{align}\label{survprobw}
p_w=\frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w} .
\end{align}

If the population originally consists of a single aneuploid cell, then the probability that the population will survive is given by the quadratic equation
\begin{align}
1-p_a=\frac{\mu_a}{\lambda_a+\mu_a+v}+\frac{v}{\lambda_a+\mu_a+v}\left(1-p_m\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v}\left(1-p_a\right)^2,
\end{align}
whose smallest solution is the survival probability:
\begin{align}\label{survproba}
p_a=\frac{\lambda_a-\mu_a-v+\sqrt{\left(\lambda_a-\mu_a-v\right)^2+4\lambda_avp_m}}{2\lambda_a}.
\end{align}

The probability that a single mutant cell survives is given by (Appendix \ref{AppendixSurvLin}):
\begin{equation}
p_m = \left\{
  \begin{array}{@{}ll@{}}
  \frac{\lambda_m-\mu_m}{\lambda_m},\quad \text{if }\lambda_m>\mu_m \\
   0,\quad \text{else} .
  \end{array}\right .
\end{equation}

An alternative method to obtain $p_m$ would be to solve the following quadratic equation:
\begin{align}
1-p_m=\frac{\mu_m}{\lambda_m+\mu_m}+\frac{\lambda_m}{\lambda_m+\mu_m}\left(1-p_m\right)^2.
\end{align}
% TODO add subsection on the Gillespie simulations OK
\subsection*{Evolutionary simulation} 
Simulations are performed using a \emph{Gillespie algorithm} \cite{gillespie1976general,gillespie1977exact} implemented in Python \cite{python}.
The simulation monitors the number of cells of each type: wildtype, aneuploid, and mutant. 
The wildtype population initially consists of $w_0$ cells, whereas the other cell types are initially absent.
\begin{table}
\begin{center}
  \begin{tabular}{| l |c| c | c | c |}
    \hline
    Symbol & Name & Value & Units & References \\ \hline
    $N$ & Tumor size at beginning of treatment& $10^7-10^9$ & cells  & \cite{del2009does} \\ \hline
    $\lambda_w$ & Wildtype birth rate& 0.14 & 1/days  & \cite{bozic2013evolutionary} \\ \hline
    $\mu_w$ & Wildtype death rate& 0.17 & 1/days  & \cite{bozic2013evolutionary} \\ \hline
    $\lambda_a$ & Aneuploid birth rate& 0.14 & 1/days  & * \\ \hline
    $\mu_a$ & Aneuploid death rate& $0.13-0.17$ & 1/days  & * \\ \hline
    $\lambda_m$ & Mutant birth rate& 0.14 & 1/days  & \cite{bozic2013evolutionary} \\ \hline
    $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \cite{bozic2013evolutionary} \\ \hline
    $u$ & Chromosomal instability rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \cite{nowak2004evolutionary,bakker2023predicting} \\ \hline
    $v$ & Mutation rate& $10^{-7}-10^{-9}$ &  1$\slash$gene$\slash$cell division  & \cite{nowak2004evolutionary} \\ \hline
  \end{tabular}
\caption{\label{demo-table}\textbf{Model parameters.} Table of parameters used in our model (see Figure \ref{figureAneuploidy}). The * symbol in the reference column means that for those parameters the values have not been selected from a paper. For the aneuploid birth rate we have chosen the same value for the wildtype and mutant birth rates. For the anueploid death rate we have chosen an intermediate value between the wildtype and mutant death rates.}
  \label{table1}
\end{center}
\end{table}

At each iteration of the simulation loop we compute the rate $\nu_j$ of each event $j$. % TODO explains what events can occur and how you compute their rates, refer to illustration figure  OK
The sate of the stochastic system at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$ which can change in the time interval $\Delta_t$ according to the following events with appropriate rates (see Figure \ref{figureAneuploidy}):
\begin{subequations}
\begin{flalign*}
(+1,0,0)&:\quad \lambda_ww_t\quad\left(\text{birth euploid cell}\right),\\
(-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death euploid cell}\right),\\
(-1,+1,0)&:\quad uw_t\quad\left(\text{euploid cell aquires aneuploidy}\right),\\
(-1,0,+1)&:\quad vw_t\quad\left(\text{euploid cell aquires mutation}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\quad\left(\text{birth aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death aneuploid cell}\right),\\
(0,-1,+1)&:\quad va_t\quad\left(\text{aneuploid cell aquires mutation}\right),\\
(0,0,+1)&:\quad \lambda_am_t\quad\left(\text{birth mutant cell}\right),\\
(0,0,-1)&:\quad \mu_am_t\quad\left(\text{death mutant cell}\right).
\end{flalign*}
\end{subequations}
We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \textit{Exp}(\sum_j \nu_j)$.
Then, we randomly determine which event occurred, where the probability for event $j$ is $p_j=\nu_j/\sum_i \nu_i$.
Finally, we update the number of cells of each type according to the event that occurred and update the time from $t$ to $t+\Delta t$. % TODO more details - you need to specify the events and how they change the number of cells. OK
We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$. % TODO how do you determine the extinction prob? ref to equation OK
We continue running the simulation until:
\begin{equation*}
m_t>\left\lfloor-\frac{3\log10}{\log\left(\frac{\mu_m}{\lambda_m}\right)}\right\rfloor+1,
\end{equation*}
which is the population size at which the extinction probability of the population of mutant cell is less then $0.1\%$. 
% TODO do you use tau leap? if so, update the text accordingly. OK
When the size of the initial population is large we utilize $\tau$-leaping where the increment of population change in a fixed time interval $\Delta t$ is Poisson distributed with mean $\nu_i\Delta t$ for each population $i$ \cite{gillespie2001approximate}. If the increment is negative and larger then the subpopulation size then updated to be zero.
\paragraph{Data availability.} All source code is available online at \url{https://github.com/yoavram-lab/EvolutionaryRescue}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TODO combine all diagram figures to a single figure with multiple panels. and make it fit in a single page. OK

\begin{figure}
\centering
\includegraphics[width=0.75\textwidth]{Figures/figureAneuploidy.pdf}
%\caption{}

\caption{
A population of cancer cells is subdivided to wildtype euploid, wildtype aneuploid and mutant euploid cells, which divide with rates $\lambda_w$, $\lambda_a$, and $\lambda_m$, respectively, and die at rates $\mu_w$, $\mu_a$, and $\mu_m$, respectively. 
Wildtype cells can become aneuploid at rate $u$. Both aneuploid and wildtype cells can acquire a beneficial mutation with rate $v$.}
\label{figureAneuploidy}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% OVERVIEW OF RESULTS

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TODO give a descriptive name instead of a mathematical formula OK
%\subsection*{First case: $4\lambda_avp_m<\left(\lambda_a-\mu_a-v\right)^2$}
\subsection*{First case: Highly deleterious or highly beneficial aneuploidy}

\paragraph{Aneuploid cells are drug resistant.} 
We first assume wildtype cells are susceptible to the drug, $\lambda_w<\mu_w$, whereas aneuploid cells are resistant, $\lambda_a>\mu_a$.
Thus, we rewrite \cref{survprobw,survproba} as
\begin{align*}
p_a&=\frac{\lambda_a-\mu_a-v}{2\lambda_a}\left(1+\sqrt{1+\frac{4\lambda_avp_m}{\left(\lambda_a-\mu_a-v\right)^2}}\right),\\
p_w&=\frac{\lambda_w-\mu_w-u-v}{2\lambda_w}\left(1-\sqrt{1+\frac{4\lambda_w\left(vp_m+up_a\right)}{\left(\lambda_w-\mu_w-u-v\right)^2}}\right).
\end{align*}
Using a quadratic Taylor expansion, $\sqrt{1+x}=1+x/2+O(x^2)$, 
we obtain the following approximation for the survival probability of a population consisting of a single individual wildtype cell (assuming $u,v \ll 1$),
\begin{align}\label{survprobwinitial}
p_w 
&\approx -\frac{vp_m+up_a}{\lambda_w-\mu_w-u-v}\\
\nonumber
%&\approx-\frac{1}{\lambda_w-\mu_w-u-v}\left[\frac{v\left(\lambda_a-\mu_a-u\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a-u\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right]\\ \label{survprobw2}
&\approx-\frac{1}{\lambda_w-\mu_w}\left[\frac{u\left(\lambda_a-\mu_a\right)}{\lambda_a}+\frac{uv\left(\lambda_m-\mu_m\right)}{\lambda_m\left(\lambda_a-\mu_a\right)}+\frac{v\left(\lambda_m-\mu_m\right)}{\lambda_m}\right].
\end{align}

We write \cref{survprobw} as
\begin{equation}\label{survprobwapprox1}
p_w=-\frac{1}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right).
\end{equation}
So, given an initial population of $N$ wildtype cells, the probability that the population will survive is given by
\begin{align}\label{aneuploidyresistentfirstapprox}
p_{resc}=1-\left(1-p_w\right)^N\approx 1-\e^{-Np_w}=1-\exp\left[\frac{N}{\Delta_w}\left(\frac{u\Delta_a}{\lambda_a}+\frac{uv\Delta_m}{\lambda_m\Delta_a}+\frac{v\Delta_m}{\lambda_m}\right)\right] .
\end{align}
\Cref{SurvPlot} for $N=10^4$ and \Cref{SurvPlotLargeN} for $N=10^8$  show that the survival probability, $p_{resc}$, quickly grows as the wildtype growth rate, $\lambda_w$, increases. % Todo I dont think this figure should show lambda_w>1, as in that case the wildtype is resistant and there is no rescue scenario. OK

>>>>

\usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{tikz,pgfplots}
\usepackage{gensymb}
\usepackage{graphicx}
\usepackage{subcaption}
%\usepackage{xurl}

@@

% NatBib
%\usepackage[super,comma]{natbib}
%\usepackage[numbers,super,compress]{natbib}
\usepackage[round,colon,authoryear]{natbib}

@@

\author[1]{Remus Stana}
\author[2]{Uri Ben-David}
\author[3]{Daniel B. Weissman}
\author[1,*]{Yoav Ram}
\affil[1]{School of Zoology, Faculty of Life Sciences, Tel Aviv University, Tel Aviv, Israel}
\affil[2]{Department of Human Molecular Genetics and Biochemistry, Faculty of Medicine, Tel Aviv University, Tel Aviv, Israel}
\affil[3]{Department of Physics, Emory University, Atlanta, GA}
\affil[*]{Corresponding author: yoav@yoavram.com}

@@

\paragraph{Aneuploidy in cancer.} Chromosomal instability (CIN) is the mitotic process in which cells suffer from chromosome mis-segregation that leads to aneuploidy, where cells are characterized by structural changes of the chromosomes and copy number alterations \citep{schukken2018cin}.
Interestingly, aberrations in chromosome copy number have been shown to allow cancer cells to survive under stressful conditions such as drug therapy.
Indeed, cancer cells are often likely to be aneuploid, and aneuploidy is associated with poor patient outcomes \citep{ben2020context}.

@@

The role of chromosomal instability (CIN) in the emergence of cancer has been studied extensively in the past decades \citep{michor2005can,christine2018understanding,nowak2002role,pavelka2010dr,komarova2003mutation,zhu2018cellular}.
One hypothesis is that CIN facilitates tumor genesis by accelerating the removal of tumor suppression genes (TSG) and subsequent appearance of cancer. The deletion of tumor suppression genes can happen in two ways: two point mutations deleting both alleles of the TSG (assuming a diploid genotype), or one point mutation and one chromosomal loss event.
Initial theoretical studies have shown that aneuploidy can have a significant role in the deletion of the the tumor suppressing genes when compared to two consecutive point mutations \citep{nowak2002role,komarova2003mutation,michor2005can,komarova2008selective}.
However, when taking into account that the appearance of aneuploidy requires a mutation to trigger CIN, the probability that CIN precedes tumor genesis is highly unlikely.

\paragraph{Evolutionary rescue.} Populations adapted to a certain environment are vulnerable to environmental changes, which might cause extinction of the population. Examples of such environmental changes include climate change, invasive species or the onset of drug therapies. Adaptation is a race against time as the population size decreases in the new environment~\citep{tanaka2022surviving}. 
\emph{Evolutionary rescue} is the process where the population acquires a trait that increases fitness in the new environment such that extinction is averted. It is mathematically equivalent to the problem of crossing of fitness valley \citep{weissman2009rate,weissman2010rate}.
There are three potential ways for a population to survive environmental change: migration to a new habitat similar to the one before the onset of environmental change \citep{cobbold2020should}; adaptation by phenotypic plasticity without genetic modification \citep{carja2019evolutionary,carja2017evolutionary,levien2021non}; and adaptation through genetic modifications, e.g., mutation \citep{uecker2014evolutionary,uecker2016role,uecker2011fixation}.

Models of evolutionary rescue usually assume that the fitness of the wildtype and mutant are homogeneous in time. An exception was given by \citet{marrec2020adapt}, who modeled the fitness of the wildtype and mutant as time dependent. Additionally, \citet{uecker2011fixation} investigated the probability of fixation of a beneficial mutation in a variable environment with arbitrary time-dependent selection coefficient and population size.
Most models focus on the probability that at least one mutation rescues the population. How multiple mutations contribute to the survival of the population is less explored, but \citet{wilson2017soft} have shown that evolutionary rescue is significantly enhanced by soft selective sweeps when multiple mutations contribute. 
Evolutionary rescue that requires two successive mutations has been investigated using diffusion approximation by \citet{martin2013probability}.

@@

We follow the number of cancer cells that have one of three different genotypes at time $t$: wildtype, $w_t$; aneuploid, $a_t$; and mutant, $m_t$. 
These cells divide and die with rates $\lambda_k$ and $\mu_k$ (for $k=w, a, m$).
The difference between the division and death rate is $\Delta_k = \lambda_k-\mu_k$.
We assume the population of cells is under a strong stress, such as drug therapy, to which the wildtype genotype is susceptible and therefore $\Delta_w<0$, whereas the mutant is resistant to the stress, $\Delta_m>0$
We analyze two cases: in the first, aneuploid cells are partially resistant, $\Delta_m>\Delta_a>0$; in the second, aneuploid cells are tolerant, $0>\Delta_a>\Delta_w$ \citep[see][for the distinction between susceptible, resistant, and tolerant]{brauner2016distinguishing}.
Wildtype cells may missegregate to become aneuploids at rate $u$. Both aneuploid and wildtype cells may mutate to become mutants at rate $v$ (\Cref{figureAneuploidy}). 
%Thus, the changes in the number of each cell type is described by 
%\begin{equation}
%\begin{aligned}
%w_t&\rightarrow w_t+1:\quad \lambda_ww_t,\\
%w_t&\rightarrow w_t-1:\quad \left(\mu_w+u+v\right)w_t,\\
%a_t&\rightarrow a_t+1:\quad \lambda_aa_t+uw_t,\\
%a_t&\rightarrow a_t-1:\quad \left(\mu_a+v\right)a_t,\\
%m_t&\rightarrow m_t+1:\quad \lambda_am_t+va_t+vm_t,\\
%m_t&\rightarrow m_t-1:\quad \mu_am_t.
%\end{aligned}
%\end{equation}

%%%%%%%%%%%
\begin{figure}[!t]
\centering
\includegraphics[width=0.5\textwidth]{Figures/figureAneuploidy.pdf}
\caption{
\textbf{Model illustration.}
% TODO in the figure change labels to Wildtype, Aneuploid, Mutant (instead of Euploid etc)
A population of cancer cells is composed of wildtype, aneuploid, and mutant cells, which divide with rates $\lambda_w$, $\lambda_a$, and $\lambda_m$ and die at rates $\mu_w$, $\mu_a$, and $\mu_m$, respectively. 
Wildtype cells can become aneuploid at rate $u$. Both aneuploid and wildtype cells can acquire a beneficial mutation with rate $v$. Color denotes the relative growth rates of the three genotypes such that $\lambda_w - \mu_w < \lambda_a - \mu_a < \lambda_m - \mu_m$.
}
\label{figureAneuploidy}
\end{figure}
%%%%%%%%%%%

Evolutionary rescue occurs when mutant cells appear and fixate ($m_t>>1$) in the population before the population becomes extinct ($w_t=a_t=m_t=0$).
Aneuploidy may contribute to evolutionary rescue by either preventing (when $\Delta_a>0$) or delaying (when $0>\Delta_a>\Delta_w$) the extinction of the population before mutant cells appear and fixate.
To analyze evolutionary rescue in this model, we use the framework of \emph{multitype branching processes} \citep{rybnikov2021fitness,harris1963theory}. 
This allows us to find explicit expressions for the \emph{survival probability}: the probability that the population does not become extinct.
Note that branching processes assume that lineages produced by cells from the initial population divide and die independently of each other, which may be unrealistic, as cells usually compete for resources.
A more realistic model may include competition for limited resources and spatial structure, which may play an important role in the development of cancer \citep[e.g.,][]{martens2011spatial}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Evolutionary simulation} 
Simulations are performed using a \emph{Gillespie algorithm} \citep{gillespie1976general,gillespie1977exact} implemented in Python \citep{python}.
The simulation monitors the number of cells of each type: wildtype, aneuploid, and mutant. 
The wildtype population initially consists of $w_0$ cells, whereas the other cell types are initially absent.

\begin{table}
\begin{center}
  \begin{tabular}{| l |p{5cm}| c | c | p{3cm} |}
    \hline
     & Name & Value & Units & References \\ \hline
    $N$ & Tumor size at beginning of treatment& $10^7-10^9$ & cells  & \citet{del2009does} \\ \hline
    $\lambda_w$ & Wildtype birth rate& 0.14 & 1/days  & \citep{bozic2013evolutionary} \\ \hline
    $\mu_w$ & Wildtype death rate& 0.17 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $\lambda_a$  & Aneuploid birth rate$^\ast$ & 0.14 & 1/days  & - \\ \hline
    $\mu_a$ & Aneuploid death rate$^\ast$ & $0.13-0.17$ & 1/days  & - \\ \hline
    $\lambda_m$ & Mutant birth rate& 0.14 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $\mu_m$ & Mutant death rate& 0.13 & 1/days  & \citet{bozic2013evolutionary} \\ \hline
    $u$ & Chromosomal instability rate& $10^{-3}-10^{-2}$ & 1$\slash$cell division  & \citet{nowak2004evolutionary,bakker2023predicting} \\ \hline
    $v$ & Mutation rate& $10^{-7}-10^{-9}$ &  1$\slash$gene$\slash$cell division  & \citet{nowak2004evolutionary} \\ \hline
  \end{tabular}
\caption{\textbf{Model parameters.} 
Aneuploid birth rate $\lambda_a$ is set to the same value as the wildtype and mutant birth rates, $\lambda_w$ and $\lambda_m$.
Aneuploid death rate $\mu_a$ is set to an intermediate value between the wildtype and mutant death rates, $\mu_w$ and $\mu_m$.}
  \label{table1}
\end{center}
\end{table}

The state of the stochastic system at time $t$ is represented by the triplet $\left(w_t,a_t,m_t\right)$. The following describes the events that may occur (right column), the rates at which they occur (middle column), and the effect these events have on the state (\Cref{figureAneuploidy}):
\begin{subequations}
\begin{flalign*}
(+1,0,0)&:\quad \lambda_ww_t\quad\left(\text{birth of wildtype cell}\right),\\
(-1,0,0)&:\quad \mu_ww_t\quad\left(\text{death of wildtype cell}\right),\\
(-1,+1,0)&:\quad uw_t\quad\left(\text{wildtype cell becomes aneuploid}\right),\\
(-1,0,+1)&:\quad vw_t\quad\left(\text{wildtype cell becomes mutant}\right),\\
(0,+1,0)&:\quad \lambda_aa_t\quad\left(\text{birth of aneuploid cell}\right),\\
(0,-1,0)&:\quad \mu_aa_t\quad\left(\text{death of aneuploid cell}\right),\\
(0,-1,+1)&:\quad va_t\quad\left(\text{aneuploid cell becomes mutant}\right),\\
(0,0,+1)&:\quad \lambda_am_t\quad\left(\text{birth of mutant cell}\right),\\
(0,0,-1)&:\quad \mu_am_t\quad\left(\text{death of mutant cell}\right).
\end{flalign*}
\end{subequations}
Each iteration of the simulation loop starts with computing the rates $\nu_j$ of each event $j$.
We then draw the time until the next event, $\Delta t$, from an exponential distribution whose rate parameter is the sum of the rates of all events, such that $\Delta t \sim \textit{Exp}(\sum_j \nu_j)$.
Then, we randomly determine which event occurred, where the probability for event $j$ is $p_j=\nu_j/\sum_i \nu_i$.
Finally, we update the number of cells of each type according to the event that occurred and update the time from $t$ to $t+\Delta t$. % TODO more details - you need to specify the events and how they change the number of cells. OK
We repeat these iterations until either the population becomes extinct (the number of cells of all types is zero) or the number of mutant cells is high enough so that its extinction probability is $<0.1\%$, that is until
\begin{equation*}
m_t>\left\lfloor-\frac{3\log10}{\log\left(\frac{\mu_m}{\lambda_m}\right)}\right\rfloor+1,
\end{equation*}

\paragraph{$\tau$-leaping.}
% TODO "is large" - how large?
When the size of the initial population is large we utilize $\tau$-leaping \citep{gillespie2001approximate}, where change in number of cells of genotype $i$ in a fixed time interval $\Delta t$ is Poisson distributed with mean $\nu_i\Delta t$.
If the increment is negative and larger then the subpopulation size then updated to be zero. % TODO "and larger..."??

\paragraph{Density-dependent growth.}
To simulate birth and death rates that depend on the number of cells in the population, we transform the rates of division and death to the following:
\begin{align*}
\lambda_w' &= \lambda_w, \\
\mu_w' &= \mu_w,\\
\lambda_a' &= C_1+\left(\lambda_a-\mu_a\right)\left(1-\frac{w+a+m}{K}\right),\\ 
\mu_a' &= C_1,\\
\lambda_m' &= C_2+\left(\lambda_m-\mu_m\right)\left(1-\frac{w+a+m}{K}\right),\\ 
\mu_m' &= C_2,
\end{align*}
where $C_1, C_2>0$ are constants. % TODO what constants?

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Code and data availability.} All source code is available online at \url{https://github.com/yoavram-lab/EvolutionaryRescue}.

@@

% TODO: OVERVIEW OF RESULTS

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Survival probability}

Let $p_w$, $p_a$, and $p_m$ be the survival probabilities of a population consisting initially of single wildtype cell, aneuploid cell, or mutant cell, respectively.
The complements $1-p_w$, $1-p_a$, and $1-p_m$ are the extinction probabilities, which satisfy each its respective equation,
\begin{equation} \label{extinction_prob}
\begin{aligned}
1-p_w = &\frac{\mu_w}{\lambda_w+\mu_w+u+v} + 
		  \frac{u}{\lambda_w+\mu_w+u+v}\left(1-p_a\right) + \\
		  & \frac{\lambda_w}{\lambda_w+\mu_w+u+v}\left(1-p_w\right)^2 +
		  \frac{v}{\lambda_w+\mu_w+u+v}\left(1-p_m\right) ,\\
1-p_a = &\frac{\mu_a}{\lambda_a+\mu_a+v}+\frac{v}{\lambda_a+\mu_a+v}\left(1-p_m\right)+\frac{\lambda_a}{\lambda_a+\mu_a+v}\left(1-p_a\right)^2 ,\\
1-p_m = &\frac{\mu_m}{\lambda_m+\mu_m}+\frac{\lambda_m}{\lambda_m+\mu_m}\left(1-p_m\right)^2 .	 
\end{aligned}
\end{equation}

The survival probabilities are given by the smallest solution for each quadratic equation \citep{uecker2015adaptive}. Therefore we have
\begin{equation}\label{survival_prob}
\begin{aligned}
p_w &= \frac{\lambda_w-\mu_w-u-v+\sqrt{\left(\lambda_w-\mu_w-u-v\right)^2+4\lambda_w\left(up_a+vp_m\right)}}{2\lambda_w} ,\\
p_a &= \frac{\lambda_a-\mu_a-v+\sqrt{\left(\lambda_a-\mu_a-v\right)^2+4\lambda_avp_m}}{2\lambda_a}, \\
p_m &= \left\{
  \begin{array}{@{}ll@{}}
  \frac{\lambda_m-\mu_m}{\lambda_m},& \lambda_m \ge \mu_m \\
   0, & \lambda_m \le \mu_m . %TODO in the appendix this is 1, not 0
  \end{array}\right .  
\end{aligned}
\end{equation}
See Appendix \ref{AppendixSurvLin} for the derivation of $p_m$. %TODO why do we need Appendix 1 if we can just find p_m as solution to quadratic equation?
Note that the equation for $p_w$ depends on both $p_a$ and $p_m$, and the equation for $p_a$ depends on $p_m$.
To proceed, we can plug the solution for $p_m$ and $p_a$ into the solution for $p_w$. We perform this for three different scenarios.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection*{Scenario 1: Aneuploid cells are drug resistant}
